Using Sorghum as an outgroup to find syntenic regions in Maize: regenerating

### 2017-12-21 SubGenome Analysis.Rmd
Major structure overhall
Moved data from "data" to "data-raw", reformatted data loading to take advantage of packaged data, moved much of the "cleaning" functionality and id mapping in do.R to clean.R.

library(topGO)
library(readr)
library(readxl)
library(ggplot2)
library(tidyr)
library(dplyr)
library(VennDiagram)
library(gridExtra)
```

@@

```{r subgenome_expression_top_ten, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
gene.foldchange <- homeologFoldChanges(maize.expression.sample.avg, homeologs.pairs)
gene.foldchange$Maize1_url <- apply(gene.foldchange[,1], 2, function(x) { wrapInHREF(getGenePageURL(x), x) })
gene.foldchange$Maize2_url <- apply(gene.foldchange[,2], 2, function(x) { wrapInHREF(getGenePageURL(x), x) })
gene.foldchange[1:10,] %>% 
  select(Maize1_url, Maize2_url, Sample, Value_maize1, Value_maize2, foldChange) %>%
  knitr::kable(caption = "Top ten differentially expressed gene pairs (in log2)", align = "llrrr")

protein.foldchange <- homeologFoldChanges(maize.protein.abundance.sample.avg, homeologs.pairs)
protein.foldchange$Maize1_url <- apply(protein.foldchange[,1], 2, function(x) { wrapInHREF(getGenePageURL(x), x) })
protein.foldchange$Maize2_url <- apply(protein.foldchange[,2], 2, function(x) { wrapInHREF(getGenePageURL(x), x) })
protein.foldchange[1:10,] %>% 
  select(Maize1_url, Maize2_url, Sample, Value_maize1, Value_maize2, foldChange) %>%
  knitr::kable(caption = "Top ten differentially abundant protein pairs (in log2)", align = "llrrr")

gene.foldchange[1:50,] %>%
  inner_join(protein.foldchange[1:50,], by = c("Maize1", "Maize2", "Sample")) %>%
  rename(foldChange_expr=foldChange.x, foldChange_prot=foldChange.y, Maize1_url=Maize1_url.x, Maize2_url=Maize2_url.x) %>%
  select(Maize1_url, Maize2_url, Sample, foldChange_expr, foldChange_prot) %>%
  knitr::kable(caption = "Gene pairs in the top 50 for both differential gene expression and differential protein abundance (in log2).", align = "llrrrr")
```

@@

```{r subgenome_expression_scatter, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=8, fig.height=8, fig.align='center', fig.cap = cap}
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.expression.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.expression.sample.avg, by=c("Maize2"="geneID", "Sample"="Sample")) %>%
  subset(!is.na(FPKM_avg.x) & !is.na(FPKM_avg.y))
names(data)[4] <- "Value_maize1"
names(data)[5] <- "Value_maize2"
# data <- subset(data, Sample %in% c("Silk", "Embryo 20 DAP", "Leaf Zone 2 (Stomatal)"))

plot <-
  ggplot(data, aes(x=Value_maize1, y=Value_maize2)) +
  geom_point(alpha=.1, aes(color = Sample)) +
  geom_abline(mapping = null, data = null, slope = 1, intercept = 0) +
  scale_x_log10() + scale_y_log10() +
  labs(
    title = paste0("Scatterplot of expression for gene pair(s).\nAlpha=.1, reference line"),
    x = "Expression for Maize1 Gene (log scale)",
    y = "Expression for Maize2 Gene (log scale)"
  ) +
  theme(legend.position = "none") + facet_wrap(~Sample, nrow=6)
plot(plot)

@@

```{r subgenome_protein_scatter, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=8, fig.height=8, fig.align='center', fig.cap = cap}
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize2"="geneID", "Sample"="Sample")) %>%
  subset(!is.na(dNSAF_avg.x) & !is.na(dNSAF_avg.y))
names(data)[4] <- "Value_maize1"
names(data)[5] <- "Value_maize2"
# data <- subset(data, Sample %in% c("Silk", "Embryo 20 DAP", "Leaf Zone 2 (Stomatal)"))

plot <-
  ggplot(data, aes(x=Value_maize1, y=Value_maize2)) +
  geom_point(alpha=.1, aes(color = Sample)) +
  geom_abline(mapping = null, data = null, slope = 1, intercept = 0) +
  scale_x_log10() + scale_y_log10() +
  labs(
    title = paste0("Scatterplot of protein abundance for gene pair(s).\nAlpha=.1, reference line"),
    x = "Abundance for Maize1 Gene (log scale)",
    y = "Abundance for Maize2 Gene (log scale)"
  ) +
  theme(legend.position = "none") + facet_wrap(~Sample, nrow=6)
plot(plot)

@@

```{r subgenome_expression_protein_foldchange_scatter, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=6, fig.height=5, fig.align='center', fig.cap = cap}
## Start with the paired expression data
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.expression.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.expression.sample.avg, by=c("Maize2"="geneID", "Sample"="Sample")) %>%
  subset(!is.na(FPKM_avg.x) & !is.na(FPKM_avg.y))
names(data)[4] <- "FPKM_maize1"
names(data)[5] <- "FPKM_maize2"

## Add paired abundance data
data <-
  data %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize1"="geneID", "Sample"="Sample")) %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize2"="geneID", "Sample"="Sample")) %>%
  subset(!is.na(dNSAF_avg.x) & !is.na(dNSAF_avg.y))
names(data)[6] <- "dNSAF_maize1"
names(data)[7] <- "dNSAF_maize2"

data$foldChange_expr <- log2(data$FPKM_maize1) - log2(data$FPKM_maize2)
data$foldChange_abun <- log2(data$dNSAF_maize1) - log2(data$dNSAF_maize2)
ggplot(data, aes(foldChange_expr, foldChange_abun)) +
  geom_point(alpha=.1) +
  xlim(-10,10) + ylim(-10,10) +
  geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95) +
  # facet_wrap(~Sample, nrow=6) +
  labs(
    title = paste0("Scatterplot of Foldchange in Expression in Maize1 vs.\nFoldchange in Abundance in Maize1\nAlpha=.1, linear fit, n=13413"),
    x = "log2(Expression Maize1) - log2(Expression Maize2)",
    y = "log2(Abundance Maize1) - log2(Abundance Maize2)"
  ) + 
  annotate("text",x=7,y=-8, label=paste("Pearson Cor = ", round(cor(data$foldChange_expr, data$foldChange_abun, method = c("pearson")),2)))

@@

```{r subgenome_expression_protein_scatter, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=8, fig.height=8, fig.align='center', fig.cap = cap}
## Start with the paired expression data
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.expression.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.expression.sample.avg, by=c("Maize2"="geneID", "Sample"="Sample")) %>%
  subset(!is.na(FPKM_avg.x) & !is.na(FPKM_avg.y))
names(data)[4] <- "FPKM_maize1"
names(data)[5] <- "FPKM_maize2"

## Add paired abundance data
data <-
  data %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize1"="geneID", "Sample"="Sample")) %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize2"="geneID", "Sample"="Sample")) %>%
  subset(!is.na(dNSAF_avg.x) & !is.na(dNSAF_avg.y))
names(data)[6] <- "dNSAF_maize1"
names(data)[7] <- "dNSAF_maize2"

## View as scatter
ggplot(data, aes(FPKM_maize1, dNSAF_maize1)) +
  geom_point(alpha=.1) +
  geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95) +
  # geom_smooth(method="loess", se=TRUE, fullrange=FALSE, level=0.95) +
  geom_abline(mapping = null, data = null, slope = 1, intercept = 0) +
  scale_x_log10() + scale_y_log10() +
  facet_wrap(~Sample, nrow=6) +
  labs(
    title = paste0("Scatterplot of Expression in Maize1 vs. Abundance in Maize1\nLog-Log Scale for gene pair(s).\nAlpha=.1, reference line (black), lm fit (blue)"),
    x = "Expression for Maize1 Gene (log scale)",
    y = "Abundance for Maize1 Gene (log scale)"
  )

@@

```{r include=FALSE}
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.expression.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.expression.sample.avg, by=c("Maize2"="geneID", "Sample"="Sample")) %>%
  subset(!is.na(FPKM_avg.x) & !is.na(FPKM_avg.y))
names(data)[4] <- "FPKM_maize1"
names(data)[5] <- "FPKM_maize2"

## Add paired abundance data
data <-
  data %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize1"="geneID", "Sample"="Sample")) %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize2"="geneID", "Sample"="Sample")) %>%
  subset(!is.na(dNSAF_avg.x) & !is.na(dNSAF_avg.y))
names(data)[6] <- "dNSAF_maize1"
names(data)[7] <- "dNSAF_maize2"

@@

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./data/Images/High expressing genes tend to vary.png')
```

@@

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./data/Images/Low Expressing Genes tend to stay low.png')
```

@@

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./data/Images/Pollen is different.png')
```

@@

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./data/Images/High-low variation.png')
```

@@

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./data/Images/Top fold-change genes.png')
```

>>>>

library(ggplot2)
library(tidyr)
library(dplyr)
library(VennDiagram)
library(gridExtra)
```

@@

```{r subgenome_expression_top_ten, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
gene.foldchange <- homeologFoldChanges(maize.expression.sample.avg, homeologs.pairs)
gene.foldchange$Maize1_url <- apply(gene.foldchange[,1], 2, function(x) { wrapInHREF(getGenePageURL(x), x) })
gene.foldchange$Maize2_url <- apply(gene.foldchange[,2], 2, function(x) { wrapInHREF(getGenePageURL(x), x) })
gene.foldchange[1:10,] %>% 
  select(Maize1_url, Maize2_url, sample, Value_maize1, Value_maize2, foldChange) %>%
  knitr::kable(caption = "Top ten differentially expressed gene pairs (in log2)", align = "llrrr")

protein.foldchange <- homeologFoldChanges(maize.protein.abundance.sample.avg, homeologs.pairs)
protein.foldchange$Maize1_url <- apply(protein.foldchange[,1], 2, function(x) { wrapInHREF(getGenePageURL(x), x) })
protein.foldchange$Maize2_url <- apply(protein.foldchange[,2], 2, function(x) { wrapInHREF(getGenePageURL(x), x) })
protein.foldchange[1:10,] %>% 
  select(Maize1_url, Maize2_url, sample, Value_maize1, Value_maize2, foldChange) %>%
  knitr::kable(caption = "Top ten differentially abundant protein pairs (in log2)", align = "llrrr")

gene.foldchange[1:50,] %>%
  inner_join(protein.foldchange[1:50,], by = c("Maize1", "Maize2", "sample")) %>%
  rename(foldChange_expr=foldChange.x, foldChange_prot=foldChange.y, Maize1_url=Maize1_url.x, Maize2_url=Maize2_url.x) %>%
  select(Maize1_url, Maize2_url, sample, foldChange_expr, foldChange_prot) %>%
  knitr::kable(caption = "Gene pairs in the top 50 for both differential gene expression and differential protein abundance (in log2).", align = "llrrrr")
```

@@

```{r subgenome_expression_scatter, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=8, fig.height=8, fig.align='center', fig.cap = cap}
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.expression.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.expression.sample.avg, by=c("Maize2"="geneID", "sample"="sample")) %>%
  subset(!is.na(FPKM_avg.x) & !is.na(FPKM_avg.y))
names(data)[4] <- "Value_maize1"
names(data)[5] <- "Value_maize2"
# data <- subset(data, sample %in% c("Silk", "Embryo 20 DAP", "Leaf Zone 2 (Stomatal)"))

plot <-
  ggplot(data, aes(x=Value_maize1, y=Value_maize2)) +
  geom_point(alpha=.1, aes(color = sample)) +
  geom_abline(mapping = null, data = null, slope = 1, intercept = 0) +
  scale_x_log10() + scale_y_log10() +
  labs(
    title = paste0("Scatterplot of expression for gene pair(s).\nAlpha=.1, reference line"),
    x = "Expression for Maize1 Gene (log scale)",
    y = "Expression for Maize2 Gene (log scale)"
  ) +
  theme(legend.position = "none") + facet_wrap(~sample, nrow=6)
plot(plot)

@@

```{r subgenome_protein_scatter, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=8, fig.height=8, fig.align='center', fig.cap = cap}
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize2"="geneID", "sample"="sample")) %>%
  subset(!is.na(dNSAF_avg.x) & !is.na(dNSAF_avg.y))
names(data)[4] <- "Value_maize1"
names(data)[5] <- "Value_maize2"
# data <- subset(data, sample %in% c("Silk", "Embryo 20 DAP", "Leaf Zone 2 (Stomatal)"))

plot <-
  ggplot(data, aes(x=Value_maize1, y=Value_maize2)) +
  geom_point(alpha=.1, aes(color = sample)) +
  geom_abline(mapping = null, data = null, slope = 1, intercept = 0) +
  scale_x_log10() + scale_y_log10() +
  labs(
    title = paste0("Scatterplot of protein abundance for gene pair(s).\nAlpha=.1, reference line"),
    x = "Abundance for Maize1 Gene (log scale)",
    y = "Abundance for Maize2 Gene (log scale)"
  ) +
  theme(legend.position = "none") + facet_wrap(~sample, nrow=6)
plot(plot)

@@

```{r subgenome_expression_protein_foldchange_scatter, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=6, fig.height=5, fig.align='center', fig.cap = cap}
## Start with the paired expression data
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.expression.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.expression.sample.avg, by=c("Maize2"="geneID", "sample"="sample")) %>%
  subset(!is.na(FPKM_avg.x) & !is.na(FPKM_avg.y))
names(data)[4] <- "FPKM_maize1"
names(data)[5] <- "FPKM_maize2"

## Add paired abundance data
data <-
  data %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize1"="geneID", "sample"="sample")) %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize2"="geneID", "sample"="sample")) %>%
  subset(!is.na(dNSAF_avg.x) & !is.na(dNSAF_avg.y))
names(data)[6] <- "dNSAF_maize1"
names(data)[7] <- "dNSAF_maize2"

data$foldChange_expr <- log2(data$FPKM_maize1) - log2(data$FPKM_maize2)
data$foldChange_abun <- log2(data$dNSAF_maize1) - log2(data$dNSAF_maize2)
ggplot(data, aes(foldChange_expr, foldChange_abun)) +
  geom_point(alpha=.1) +
  xlim(-10,10) + ylim(-10,10) +
  geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95) +
  # facet_wrap(~sample, nrow=6) +
  labs(
    title = paste0("Scatterplot of Foldchange in Expression in Maize1 vs.\nFoldchange in Abundance in Maize1\nAlpha=.1, linear fit, n=13413"),
    x = "log2(Expression Maize1) - log2(Expression Maize2)",
    y = "log2(Abundance Maize1) - log2(Abundance Maize2)"
  ) + 
  annotate("text",x=7,y=-8, label=paste("Pearson Cor = ", round(cor(data$foldChange_expr, data$foldChange_abun, method = c("pearson")),2)))

@@

```{r subgenome_expression_protein_scatter, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=8, fig.height=8, fig.align='center', fig.cap = cap}
## Start with the paired expression data
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.expression.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.expression.sample.avg, by=c("Maize2"="geneID", "sample"="sample")) %>%
  subset(!is.na(FPKM_avg.x) & !is.na(FPKM_avg.y))
names(data)[4] <- "FPKM_maize1"
names(data)[5] <- "FPKM_maize2"

## Add paired abundance data
data <-
  data %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize1"="geneID", "sample"="sample")) %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize2"="geneID", "sample"="sample")) %>%
  subset(!is.na(dNSAF_avg.x) & !is.na(dNSAF_avg.y))
names(data)[6] <- "dNSAF_maize1"
names(data)[7] <- "dNSAF_maize2"

## View as scatter
ggplot(data, aes(FPKM_maize1, dNSAF_maize1)) +
  geom_point(alpha=.1) +
  geom_smooth(method="lm", se=TRUE, fullrange=FALSE, level=0.95) +
  # geom_smooth(method="loess", se=TRUE, fullrange=FALSE, level=0.95) +
  geom_abline(mapping = null, data = null, slope = 1, intercept = 0) +
  scale_x_log10() + scale_y_log10() +
  facet_wrap(~sample, nrow=6) +
  labs(
    title = paste0("Scatterplot of Expression in Maize1 vs. Abundance in Maize1\nLog-Log Scale for gene pair(s).\nAlpha=.1, reference line (black), lm fit (blue)"),
    x = "Expression for Maize1 Gene (log scale)",
    y = "Abundance for Maize1 Gene (log scale)"
  )

@@

```{r include=FALSE}
data <-
  homeologs.pairs %>%
  subset(Maize1 != "" & Maize2 != "") %>%
  select(Maize1, Maize2) %>%
  distinct() %>%
  inner_join(maize.expression.sample.avg, by=c("Maize1"="geneID")) %>%
  inner_join(maize.expression.sample.avg, by=c("Maize2"="geneID", "sample"="sample")) %>%
  subset(!is.na(FPKM_avg.x) & !is.na(FPKM_avg.y))
names(data)[4] <- "FPKM_maize1"
names(data)[5] <- "FPKM_maize2"

## Add paired abundance data
data <-
  data %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize1"="geneID", "sample"="sample")) %>%
  inner_join(maize.protein.abundance.sample.avg, by=c("Maize2"="geneID", "sample"="sample")) %>%
  subset(!is.na(dNSAF_avg.x) & !is.na(dNSAF_avg.y))
names(data)[6] <- "dNSAF_maize1"
names(data)[7] <- "dNSAF_maize2"

@@

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./data-raw/Images/High expressing genes tend to vary.png')
```

@@

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./data-raw/Images/Low Expressing Genes tend to stay low.png')
```

@@

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./data-raw/Images/Pollen is different.png')
```

@@

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./data-raw/Images/High-low variation.png')
```

@@

```{r echo=FALSE, out.width='100%'}
knitr::include_graphics('./data-raw/Images/Top fold-change genes.png')
```
